The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-11-17 theoretical-consensus/article/article.tex
add page 10

\author{Eduardo Ribas Brito \\ eduardo.ribas.brito@ut.ee}

@@

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by first generating the agreeable value, 
in this particular case, a block and its proof, disseminating the information 
to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-authority mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

These aim at achieving deterministic finality, by employing a BFT
consensus protocol, not at the block proposal level, but at the
block acceptance level. The idea is to have a free flow of blocks
proposed by the network, but to have a quorum of entities that
are responsible for the block acceptance. The block proposal mechanism
can be of any kind, for example, Chain-based \emph{PoS}, or Committee-based \emph{PoS},
or even pure Proof-of-Work. The block acceptance layer then allows
for a deterministic acceptance of the blocks, by having a sort of permissioned
BFT consensus protocol, with the trivial 1/3 Byzantine tolerance. The most well known examples
and variations are the Tendermint \cite{buchman2016tendermint} and the new Ethereum Casper 
\cite{buterin2017casper} protocols.

@@

\subsubsection{Proof-of-Authority} is another well known derivative,
which, in effect, is a form of Proof-of-Stake, where the stake is the 
identity of the entity, that has been verified, scrutinized and
approved by the network, giving it the right to participate in the
closed consensus group. The consensus protocol, if not a BFT protocol, 
can follow the style of a Proof-of-Work protocol, depending on the 
requirements of the network - a choice that may shape the fault tolerance,
decentralization, security and performance of the network \cite{survey-dist-consensus}.

@@

\subsubsection{Proof-of-Location} may be an alternative for envisioning a future where the
consensus is achieved by taking advantage of the dynamism and easiness of movement, placement and 
tracking of smaller and more location-versatile devices, such as mobile phones, IoT devices, etc.
The idea behind this approach is to use the location and movement of these devices 
as a proof for the consensus. It relies on the location validation by close proximity, 
for example, using low-range communication channels, either by proving the location
of the device at a certain point in time, or by proving their proximity to other devices.
Incentive mechanisms may play an important role to prevent location spoofing, collusion, etc. \cite{natoli2019deconstructing}.
These protocols are getting increased attention, especially in the context of IoT, 
with the provision of decentralized, permissionless and autonomous networks of 
low-powered devices that can be used for providing location verification services \cite{9376868}.

@@

\subsubsection{DAG structures} may diverge from the previous approaches with an envisioned 
non-linear structure for the result of the consensus attestations. The choice resides in 
directed acyclic graphs that should expand and physically connect, further in time, the new
data appended to the network. The data, which is typically held in the vertices of the graph, 
can have more or less granularity, either by storing blocks of transactions, or raw transactions directly.
This design decision may pose new potential challenges, related to the verification and finalization of
the consensus protocol. Currently, those two approaches are then distinguished as \emph{blockDAG} and 
\emph{txDAG} \cite{survey-dist-consensus}. 
The block-based DAG approach diverges from the traditional blockchain structure by flexibly 
connecting every new block to multiple previous blocks. The rest of the protocol may work in the same way 
as the traditional Proof-of-Work, however, this new feature yields new security considerations. These protocols
are theoretically designed for very high throughput in permissionless environments, but the security trade-offs
are still not well-known \cite{sompolinsky2018phantom}.
In a transaction-based DAG, every vertex holds one transaction, that is connected to previous vertices. 
All the sibling vertices shall then represent disjoint transactions to successfully deal with history and conflict
resolutions. Two well-known examples of this approach are IOTA Tangle \cite{popov2018tangle} 
and Nano \cite{lemahieu2018nano}.
The former is of particular interest, as it is especially designed for IoT applications, low-powered devices, and
micro-transactions. The protocol states that every new vertex needs to be connected to two previous vertices, 
therefore approving the parent transactions, as an incentive mechanism for honest behavior \cite{survey-dist-consensus}. 
The users are the ones to submit transactions, generating and attaching a \emph{PoW} proof to it and 
so guaranteeing probabilistic finality with 1/2 fault tolerance. This solution may indeed pair well with 
decentralization and scalability, however, there are known security issues related to the
easiness of growing malicious parallel branches with little computational effort, and those issues have been 
addressed by the introduction of an undesirable centralized solution. 
Other interesting and worth mentioning protocols are the ones that combine the DAG structure with the BFT consensus, 
such as the Snowflake-Avalanche protocol \cite{rocket2018snowflake}.

@@

\begin{itemize}
  \item \emph{Scalability}:  in terms of the number of available devices - decentralizing may allow
  for better, faster and less costly maintenance of the network, assuring more availability and reliability
  and cutting the single-entity dependency;
  \item \emph{Security}: in terms of the network as a whole - removing the single-point-of-failure 
  may reduce the attack surface, and may also allow or require more flexible and dynamic security policies;
  \item \emph{Privacy}: in terms of the network data exchange features - decentralizing may allow for
  a user-centric approach, where the users are in control of the data that is shared with the network; 
\end{itemize}

@@

A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
However, there are still known security and centralization limitations, as presented in the sections before \cite{8168250}.
Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
technique that allows for the partitioning of the network into smaller sub-networks,
that are then managed by different entities, validating transactions in parallel. 
This approach may be used to overcome the scalability bottleneck,
but it may also introduce new issues that are still to be researched.
Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
However, this does not address the hypothesis of taking advantage of the very nature of these limited environments, 
as simply choosing a BFT protocol that is resource-efficient
may also put us far apart from the ultimate decentralization and scalability goal of 
other and fairer means of consensus participation.
An interesting idea is based on the use of \emph{Proof of Location} for 
coordinating the consensus. The easiness of movement and targetted location deployment
of a lot of these low-power devices may be used to achieve a decentralized consensus
protocol, where elected nodes may be somehow chosen based on their location, 
either by proximity, or by some other witnessing mechanism. Variations and 
extensions of this idea, with different and wide use cases, have been seen in
\cite{foam2018location, xyo2018location}

@@

In this paper, we have presented a comprehensive summary of the permissionless
consensus problem, the most widely used consensus protocols, and the challenges
that arise when applying these protocols to resource-constrained networks.
We have presented the challenges that these networks are facing, and the 
limitations of the existing consensus protocols.
We have also presented the main approaches that have been explored to address these challenges,
and we have showed potential paths for future research.

@@

@@

@@

@@

\begin{figure*}[h]
  \centering
  \captionsetup{justification=centering}
  % \includegraphics{in-browser-pow-topology}
  \caption{blblb}
  \label{fig:in-browser-pow-topology}
\end{figure*}

@@

@@

The prototype was developed as a proof-of-concept and a demonstration of the feasibility
of the \emph{PoW} consensus protocol in a permissionless environment. The next steps
for this project would be to extend the functionality and the features of the system,
in order to make it more robust and usable. Some of the ideas for future work are, for example,
the improvement of the \emph{PoW} algorithm, by adding a difficulty adjustment mechanism,
to make the mining process more fair for computationally weaker devices, or to
add a mechanism for the synchronization of the old chain history. Another idea would be to
improve the parallelization of the computations, by more efficiently distributing the
workload among mining threads. Finally, the user interface could be improved 
with more informative and interactive features, that would allow the user to have a 
deeper understanding of the system and the consensus process.

>>>>

\author{Eduardo Ribas Brito 
\\Supervisors: Ulrich Norbisrath, Eero Vainikko
\\Distributed Systems Seminar, Fall 2022
\\\emph{Institute of Computer Science}
\\\emph{University of Tartu, Estonia}
\\ eduardo.ribas.brito@ut.ee}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{building-blocks-consensus.png}
  \caption{An illustration of the permissionless consensus building blocks - 
  the block generation and proposal phases followed by the block validation, 
  along with frequent network topology changes and the consequent serialization of the information.}
  \label{fig:building-blocks-consensus}
\end{figure}

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by first generating the agreeable value, 
in this particular case, a block and its proof, disseminating the information 
to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached (See Fig. \ref{fig:building-blocks-consensus}).
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-authority mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

These aim at achieving deterministic finality, by employing a BFT
consensus protocol, not at the block proposal level, but at the
block acceptance level. The idea is to have a free flow of blocks
proposed by the network, but to have a quorum of entities that
are responsible for the block acceptance. The block generation and proposal mechanism
can be of any kind, for example, Chain-based \emph{PoS}, or Committee-based \emph{PoS},
or even pure Proof-of-Work. The block acceptance layer then allows
for a deterministic acceptance of the blocks, by having a sort of permissioned
BFT consensus protocol, with the trivial 1/3 Byzantine tolerance. The most well known examples
and variations are the Tendermint \cite{buchman2016tendermint} and the new Ethereum Casper 
\cite{buterin2017casper} protocols.

@@

\subsubsection{Proof-of-Authority} is another well known derivative
which, in effect, is a form of Proof-of-Stake, where the stake is the 
identity of the entity, that has been verified, scrutinized and
approved by the network, giving it the right to participate in the
closed consensus group. The consensus protocol, if not a BFT protocol, 
can follow the style of a Proof-of-Work protocol, depending on the 
requirements of the network - a choice that may shape the fault tolerance,
decentralization, security and performance of the network \cite{survey-dist-consensus}.

@@

\subsubsection{Proof-of-Location} may be an alternative for envisioning a future where
consensus is achieved by taking advantage of the dynamism and easiness of movement, placement and 
tracking of smaller and more location-versatile devices, such as mobile phones, IoT devices, etc.
The idea behind this approach is to use the location and movement of these devices 
as a proof for the consensus. It relies on the location validation by close proximity, 
for example, using low-range communication channels, either by proving the location
of the device at a certain point in time, or by proving their proximity to other devices.
Incentive mechanisms may play an important role to prevent location spoofing, collusion, etc. \cite{natoli2019deconstructing}.
These protocols are getting increased attention, especially in the context of IoT, 
with the provision of decentralized, permissionless and autonomous networks of 
low-powered devices that can be used for providing location verification services \cite{9376868}.

@@

\subsubsection{DAG structures} may diverge from the previous approaches with an envisioned 
non-linear structure for the result of the consensus attestations. The choice resides in 
directed acyclic graphs that should expand and physically connect, further in time, the new
data appended to the network. The data, which is typically held in the vertices of the graph, 
can have more or less granularity, either by storing blocks of transactions, or raw transactions directly.
This design decision may pose new potential challenges, related to the verification and finalization of
the consensus protocol. Currently, those two approaches are then distinguished as \emph{blockDAG} and 
\emph{txDAG} \cite{survey-dist-consensus}. 
The block-based DAG approach diverges from the traditional blockchain structure by flexibly 
connecting every new block to multiple previous blocks. The rest of the protocol may work in the same way 
as the traditional Proof-of-Work, however, this new feature yields new security considerations. These protocols
are theoretically designed for very high throughput in permissionless environments, but the security trade-offs
are still not well-known \cite{sompolinsky2018phantom}.
In a transaction-based DAG, every vertex holds one transaction, that is connected to previous vertices. 
All the sibling vertices shall then represent disjoint transactions to successfully deal with history and conflict
resolutions. Two well-known examples of this approach are IOTA Tangle \cite{popov2018tangle} 
and Nano \cite{lemahieu2018nano}.
The former is of particular interest, as it is especially designed for IoT applications, low-powered devices, and
microtransactions. The protocol states that every new vertex needs to be connected to two previous vertices, 
therefore approving the parent transactions, as an incentive mechanism for honest behavior \cite{survey-dist-consensus}. 
The users are the ones to submit transactions, generating and attaching a \emph{PoW} proof to it and 
so guaranteeing probabilistic finality with 1/2 fault tolerance. This solution may indeed pair well with 
decentralization and scalability, however, there are known security issues related to the
easiness of growing malicious parallel branches with little computational effort, and those issues have been 
addressed by the introduction of an undesirable centralized solution. 
Other interesting and worth mentioning protocols are the ones that combine the DAG structure with the BFT consensus, 
such as the Snowflake-Avalanche protocol \cite{rocket2018snowflake}.

@@

\begin{itemize}
  \item \emph{Scalability}:  in terms of the number of available devices - decentralizing may allow
  for better, faster and less costly maintenance of the network, assuring more availability and reliability
  and cutting the single-entity dependency;
  \item \emph{Security}: in terms of the network as a whole - removing the single-point-of-failure 
  may reduce the attack surface, and may also allow for more flexible and dynamic security policies;
  \item \emph{Privacy}: in terms of the network data exchange features - decentralizing may allow for
  a user-centric approach, where the users are in control of the data that is shared with the network; 
\end{itemize}

@@

A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
However, there are still known security and centralization limitations, as presented in the sections before \cite{8168250}.
Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
technique that allows for the partitioning of the network into smaller subnetworks,
that are then managed by different entities, validating transactions in parallel. 
This approach may be used to overcome the scalability bottleneck,
but it may also introduce new issues that are still to be researched.
Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
However, this does not address the hypothesis of taking advantage of the very nature of these limited environments, 
as simply choosing a BFT protocol that is resource-efficient
may also put us far apart from the ultimate decentralization and scalability goal of 
other and fairer means of consensus participation.
An interesting idea is based on the use of \emph{Proof of Location} for 
coordinating the consensus. The easiness of movement and targeted location deployment
of a lot of these low-power devices may be used to achieve a decentralized consensus
protocol, where elected nodes may be somehow chosen based on their location, 
either by proximity, or by some other witnessing mechanism. Variations and 
extensions of this idea, with different and wide use cases, have been seen in
\cite{foam2018location, xyo2018location}

@@

In this paper, it was presented a comprehensive summary of the permissionless
consensus problem, along with a short analysis of the most widely used consensus 
protocols, and the challenges that arise when applying these protocols to different environments.
In particular, this paper presented the challenges that resource-constrained networks may be facing, 
and the limitations of the existing consensus protocols when applied to this type of networks.
We have also presented the main approaches that have been explored to address these challenges,
and showed some potential paths for future research.

@@

\newpage

@@

\bigbreak

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{in-browser-pow-topology}
  \caption{The topology of the system, with an underlay network of relay servers that allow the overlay network of devices to discover their peers and establish direct connections.}
  \label{fig:in-browser-pow-topology}
\end{figure}

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{in-browser-pow}
  \caption{The user interface showing three browser tabs opened and competing for mining the next blocks, 
  while achieving consensus and presenting the same view.}
  \label{fig:in-browser-pow}
\end{figure}

@@

The prototype was developed as a proof-of-concept and a demonstration of the feasibility
of the \emph{PoW} consensus protocol in a permissionless environment. The next steps
for this project would be to extend the functionality and the features of the system,
in order to make it more robust and usable. Some ideas for future work are, for example,
the improvement of the \emph{PoW} algorithm, by adding a difficulty adjustment mechanism,
to make the mining process more fair for computationally weaker devices, or to
add a mechanism for the synchronization of the old chain history. Another idea would be to
improve the parallelization of the computations, by more efficiently distributing the
workload among mining threads. Finally, the user interface could be improved 
with more informative and interactive features, that would allow the user to have a 
deeper understanding of the system and the consensus process (See Fig. \ref{fig:in-browser-pow}).

\subsection{Repository}

All the work developed, both the practical and the theoretical components, can be found 
at \emph{\url{https://github.com/edurbrito/dist-sys-seminar}}.
